---
title: "Multi-Database Benchmarking: Old vs New Cohort Generation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Database Benchmarking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette describes how to benchmark **CDMConnector::generateCohortSet** (old,
CIRCE-based) against **CohortDAG::generateCohortSet2** (new, DAG-optimized
batch) across multiple database platforms. The benchmarking script:

1. Runs both methods on each database with the same cohort set.
2. Records **overall time** for each method and writes results to a CSV.
3. **Confirms that the two cohort tables have identical rows** (order ignored) and
   writes per-database and per-cohort equivalence results to a second CSV.

Supported platforms include **PostgreSQL**, **Redshift**, **Snowflake**, **Spark**,
and **SQL Server**. You provide a named list of CDM reference objects; the script
handles timing, comparison, and CSV output.

## Credits and status

This work relies heavily on [OHDSI/circe-be](https://github.com/OHDSI/circe-be),
by Chris Knoll and others, as well as
[OHDSI/SqlRender](https://github.com/OHDSI/SqlRender), by Martijn Schuemie and
others, and [OHDSI/circepy](https://github.com/OHDSI/circepy) by `@azimov`.
This package was generated using AI coding tools and is experimental code.

## Performance improvements with the new approach

The new approach (`generateCohortSet2`) uses a **DAG-based batch optimizer** that:

- **Shares vocabulary expansion**: Concept set expansion and codeset building are
  done once and reused across all cohorts, instead of once per cohort.
- **Shares domain scans**: Filtered domain tables (e.g. drug exposure, condition
  occurrence) are built once and read by every cohort that needs them.
- **Reduces I/O**: A single batch script writes to shared staging tables and
  finalizes in one pass, instead of N separate DELETE/INSERT cycles.

As a result, **wall-clock time typically decreases** as the number of cohorts
and the overlap in concept sets increase. The ratio (new time / old time) is
often **below 1.0**, with larger batches showing greater speedups. The exact
improvement depends on:

- Number of cohorts and size of each definition
- Overlap in concept sets and domains across cohorts
- Database engine and hardware

The benchmarking script records `time_old_sec`, `time_new_sec`, and
`ratio_new_over_old` per database so you can measure the speedup on your own
data and platforms.

## Bundled tier-2 full-library benchmark

The package also includes a heavier **tier-2** benchmark that exercises the
bundled JSON cohort library end to end. This run:

1. Unzips the packaged cohort JSON files.
2. Splits the library into batches of `50` cohorts.
3. Builds a synthetic CDM for each batch with
   `CDMConnector::cdmFromCohortSet(n = 500)`.
4. Runs both `CDMConnector::generateCohortSet()` and
   `CohortDAG::generateCohortSet2()`.
5. Compares the resulting cohort tables row by row and writes batch logs, timing
   summaries, and mismatch details.

This benchmark is implemented in
`tests/testthat/test-full-library.R` and writes artifacts to
`inst/benchmark-results/`.

### Latest tier-2 run summary

The latest full-library run in this repository was completed on **March 15,
2026** using all bundled cohorts.

```{r, echo = FALSE}
tier2_summary <- data.frame(
  metric = c(
    "Cohorts tested",
    "Batches run",
    "Batches OK",
    "Batches mismatch",
    "Batches error",
    "Overall accuracy",
    "Old total time (sec)",
    "New total time (sec)",
    "Overall speedup (old/new)",
    "Median batch speedup (old/new)",
    "Rows old",
    "Rows new"
  ),
  value = c(
    "3179",
    "64",
    "54",
    "9",
    "1",
    "99.68%",
    "603.395",
    "377.373",
    "1.599x",
    "2.012x",
    "540315",
    "540281"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(tier2_summary, col.names = c("Metric", "Value"))
```

Interpretation:

- `generateCohortSet2()` was faster overall: `377.373s` versus `603.395s`.
- End-to-end parity is not complete yet: `3119` of `3129` comparable cohorts
  matched exactly, for `99.68%` accuracy.
- One batch did not compare because `cdmFromCohortSet()` failed for both methods
  with `9 observations outside observation period`.

The remaining mismatches in that run were logged in
`inst/benchmark-results/tier2_mismatch_details_latest.csv` and came from these
batches:

- `17`
- `18`
- `23`
- `26`
- `27`
- `38`
- `47`
- `58`
- `61`

## How to run the benchmark

### Prerequisites

- **CDMConnector** and **CohortDAG** installed (or `devtools::load_all()`
  for the latter).
- One or more **live CDM connections** (e.g. Postgres, Redshift, Snowflake, Spark,
  SQL Server) as CDM reference objects.
- A **cohort set** (e.g. from `CDMConnector::readCohortSet("path/to/cohorts")`).

### Bundled full-library tier-2 benchmark

To reproduce the packaged full-library benchmark, run the tier-2 test directly:

```sh
Rscript -e 'pkgload::load_all(".", quiet = TRUE); Sys.setenv(RUN_TIER_2 = "true"); testthat::test_file("tests/testthat/test-full-library.R")'
```

The default behavior matches the current repository benchmark:

- `TIER2_BATCH_SIZE=50`
- `TIER2_N_SUBJECTS=500`
- `TIER2_DISABLE_COHD=true`
- `TIER2_N_CORES=min(4, detectCores(logical = FALSE))`
- output written to `inst/benchmark-results/`

Useful overrides:

- `TIER2_BATCH_START` and `TIER2_BATCH_END`: rerun only a slice of the library.
- `TIER2_MAX_COHORTS`: limit the number of bundled JSON files loaded.
- `TIER2_RESULTS_DIR`: write results somewhere other than
  `inst/benchmark-results/`.
- `TIER2_SEED_BASE`: change the synthetic-CDM seed sequence.
- `TIER2_N_CORES`: parallelize across batches.

Example: rerun only batches `17` through `23` into a temporary results folder.

```sh
TIER2_BATCH_START=17 \
TIER2_BATCH_END=23 \
TIER2_RESULTS_DIR=/tmp/cohortdag-tier2 \
Rscript -e 'pkgload::load_all(".", quiet = TRUE); Sys.setenv(RUN_TIER_2 = "true"); testthat::test_file("tests/testthat/test-full-library.R")'
```

The tier-2 harness writes four main artifacts:

- `tier2_summary_<timestamp>.csv`: overall totals, speed, and accuracy.
- `tier2_batch_results_<timestamp>.csv`: one row per batch with timings and
  accuracy.
- `tier2_mismatch_details_<timestamp>.csv`: one row per mismatched cohort with
  row-count deltas.
- `tier2_logs_<timestamp>/batch_XXX.log`: per-batch progress and error logs.

### Single database

For one CDM, use the single-database benchmark and optional equivalence check:

```r
source("extras/benchmark_cohort_generation.R")

cohort_set <- CDMConnector::readCohortSet("path/to/cohorts")
result <- benchmark_cohort_generation(cdm, cohort_set, cohort_path = "path/to/cohorts")

# Compare old vs new cohort tables (identical rows, order ignored)
cmp <- compare_cohort_tables(result$cdm, name_old = "cohort_bench_old", name_new = "cohort_bench_new")
cmp$identical   # TRUE if same set of rows
cmp$per_cohort  # Per-cohort row counts and match status
```

### Multiple databases

Pass a **named list of CDM objects**; names are used as the `database` identifier
in the output CSVs (e.g. `postgres`, `redshift`, `snowflake`, `spark`, `sql_server`):

```r
source("extras/benchmark_cohort_generation.R")
source("extras/benchmark_multi_database.R")

cohort_set <- CDMConnector::readCohortSet("path/to/cohorts")
cdms <- list(
  postgres   = cdm_postgres,
  redshift   = cdm_redshift,
  snowflake  = cdm_snowflake,
  spark      = cdm_spark,
  sql_server = cdm_sqlserver
)

run_benchmark_multi_database(
  cdms = cdms,
  cohort_set = cohort_set,
  cohort_path = "path/to/cohorts",
  results_csv = "benchmark_results.csv",
  equivalence_csv = "benchmark_equivalence.csv"
)
```

- **benchmark_results.csv**: one row per database with `database`, `time_old_sec`,
  `time_new_sec`, `ratio_new_over_old`, `n_cohorts`, `files_included`, `status`.
- **benchmark_equivalence.csv**: one row per database (overall) plus one row per
  (database, cohort_definition_id) with `n_old`, `n_new`, `rows_identical`, `status`.

## Benchmark results CSV (timing)

The timing CSV has one row per database. Example structure:

```{r, echo = FALSE}
results_example <- data.frame(
  database = c("postgres", "redshift", "snowflake", "sql_server"),
  time_old_sec = c(120.5, 95.2, 88.1, 110.3),
  time_new_sec = c(45.2, 38.0, 32.5, 42.1),
  ratio_new_over_old = c(0.38, 0.40, 0.37, 0.38),
  n_cohorts = 4L,
  files_included = "cohort_a.json; cohort_b.json; cohort_c.json; cohort_d.json",
  status = "ok",
  stringsAsFactors = FALSE
)
knitr::kable(results_example, digits = 2)
```

- **ratio_new_over_old** &lt; 1 means the new method was faster.
- **files_included** lists the cohort definition files (or names) in the cohort set.

## Equivalence CSV (same results)

The equivalence CSV confirms that the **old and new cohort tables contain the same
rows** (order ignored). Each database has:

1. An **overall** row (with `cohort_definition_id` NA): total row counts and whether
   the full tables match.
2. **Per-cohort** rows: row counts from the old table (`n_old`), from the new
   table (`n_new`), and whether the set of rows for that cohort is identical
   (`rows_identical`).

Example:

```{r, echo = FALSE}
equiv_example <- data.frame(
  database = c("postgres", "postgres", "postgres", "postgres", "redshift", "redshift"),
  cohort_definition_id = c(NA, 1L, 2L, 3L, NA, 1L),
  n_old = c(15000L, 5000L, 6000L, 4000L, 15000L, 5000L),
  n_new = c(15000L, 5000L, 6000L, 4000L, 15000L, 5000L),
  rows_identical = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE),
  status = "ok",
  stringsAsFactors = FALSE
)
knitr::kable(equiv_example)
```

When **rows_identical** is TRUE for all cohorts (and the overall row), the new
approach produces **exactly the same cohort membership and dates** as the old
CIRCE-based method; only execution strategy and performance differ.

## Summary

| Aspect | Description |
|--------|-------------|
| **Performance** | The new batch optimizer typically reduces wall-clock time (ratio &lt; 1) by sharing vocabulary and domain work across cohorts. |
| **Correctness** | The benchmarking pipeline compares old and new cohort tables row-by-row (order ignored) and writes equivalence results to CSV. |
| **Platforms** | Run the same cohort set on Postgres, Redshift, Snowflake, Spark, and SQL Server by passing a named list of CDMs to `run_benchmark_multi_database()`. |

Use the generated CSVs to document speedups and to confirm identical results across
databases and between the two cohort generation methods.
